---
title: "Inside nlmixr2nn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inside nlmixr2nn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
.hasRx <- requireNamespace("rxode2", quietly = TRUE) &&
  requireNamespace("nlmixr2nn", quietly = TRUE)
```

This is for people changing `nlmixr2nn`, not people using it. For the latter,
see `vignette("nlmixr2nn")`.

Most of what follows is one idea repeated: **a neural network inside an ODE has
to be plumbed through machinery that was not built for it, and almost every way
of getting that plumbing wrong produces a wrong number rather than an error.**
Four separate bugs of exactly that shape have been fixed here. Each trained,
converged, and passed the whole test suite while computing something other than
what it claimed. The invariants below are the ones that turned out to matter,
and each is named with the failure it prevents.

## The shape of it

```
nn() at parse time                          R/nnUdf.R, R/nnInit.R
  |  draws weights with R's RNG, emits nn<K>(id, ...)
  v
.nnAdopt() at model assembly                R/nnAdopt.R
  |  weights -> rxForcedPars, shapes -> sticky nnMeta
  v
solve            .nnRehydrate() binds the C registry, per solve
  or
fit              .nnRun() owns the registry for the duration
                   |- augmented model built                R/nnEst.R
                   |- inner fit (estimator's own model)
                   |- cotangent captured                   src/nlmixr2nnContrib.c
                   |- weight gradient assembled + torch step
                   v
                 trained weights -> rxForcedPars on the fit's ui
```

## Parse time: where the weights come from

`nn()` is an `rxode2::rxUdfUi()` user-defined function. It expands to a compiled
call and a declaration of the weights:

```{r, eval = .hasRx}
library(nlmixr2nn)
mod <- function() {
  model({
    d/dt(centr) <- -nn(centr, nHidden = 3) * centr
  })
}
rxode2::rxSetSeed(1)
ui <- rxode2::rxode2(mod)
cat(paste(ui$lstChr, collapse = "\n"))
```

Two decisions are visible there.

**The weights are covariates, not `param()` thetas.** Thetas get theta-expanded
and mangled during FOCEi model assembly, and they enter mu-referencing.
Covariates are contiguous in `par_ptr`, do neither, and can be supplied by the
model rather than the data. The dummy `rx_nnw0_` line exists only so rxode2's
covariate detection sees the names.

**The weights are drawn when the model is parsed**, with R's RNG, in
`R/nnInit.R`, drawn with rxode2's threefry generator (`rxnorm`/`rxunif`)
inside `rxWithSeed()`, which restores both the R stream and the rxode2 seed. So
`rxSetSeed()` reproduces a network -- and does so under threads, which R's RNG
cannot promise -- and a parsed model is
solvable immediately without libtorch. torch is loaded *from* these values
rather than generating its own, which is what makes a whole fit reproducible
under a single seed.

The draw is isolated: building a model does not advance the caller's stream.
Adding a network to a model must not shift the draws in the rest of someone's
script.

### Getting a value out of a UDF

`rxUdfUi()` can return only `replace`, `before`, `after`, `iniDf`, `uiUseData`
and `uiUseMv` — code, never a value. And once a ui is compressed,
`rxUiDecompress()` returns a *fresh environment on every call*, so an in-place
assignment made after assembly is invisible to whoever holds the model.

That is why `rxRegisterUiAssembled()` exists (added to rxode2 for this):
`.nnAdopt()` runs while the ui is still a mutable environment, and moves the
drawn weights onto it as `rxForcedPars` plus a sticky `nnMeta`.

> **Not `uiUseData`.** It is tempting to think a UDF could see the dataset at fit
> time via `uiUseData = TRUE` and scale its inputs accordingly. It cannot: that
> re-parse operates on the *already-expanded* model text, where `nn(centr)` has
> become `nn1(0, centr)`, so the UDF never runs again. Verified with a probe UDF.

## The weight-block offset: the sharpest edge here

The compiled evaluator reads its weights from a contiguous block of the solve
parameter vector, starting at an offset registered beforehand with `nnSetMeta()`.

**Every model in the pipeline puts that block somewhere different.** For one
model:

```
base ui             weights at 1..13   (base 0)
saemModel           weights at 2..14   (base 1)
foceiModel$inner    weights at 3..15   (base 2)
augmented model     different again
```

Registering the wrong one **does not fail**. The evaluator reads a different
stretch of the parameter vector and computes a different network, and the
training loop then optimizes *that* — self-consistently, and wrongly. The
symptom, when it was finally measured, was a fit whose own `g` differed from the
network at its trained weights by more than the entire range of `g`.

So:

> **Invariant.** The offset must be resolved against the model that is *actually
> about to be solved* — never reconstructed from a guess at the layout.

`.nnEstSolveParams(ui, est)` answers this per estimator, and `.nnSetNetBases()`
applies it. The FOCEi posthoc that non-FOCEi estimators use to capture their
cotangents is itself a FOCEi fit, so it needs FOCEi's offset for its duration
and the round's restored afterwards.

For an ordinary solve this is handled by the ui-prep hook, which receives the
solve model — the reason rxode2's `rxRegisterUiPrep()` gained a second argument.

## Cotangent capture

`src/nlmixr2nnContrib.c` registers a bundle with nlmixr2est whose
per-observation hook records `d(LL)/d(f)` — the exact score for whatever
residual model the user wrote. Two constraints, both learned the hard way.

> **The hook runs inside nlmixr2est's OpenMP region.** It must not allocate, must
> not call into R, and must not resize anything. It used to grow its own store
> with `R_chk_realloc()` from a worker thread; that is a data race on the buffer,
> a use-after-free for every other thread mid-write, and an R API call off the
> main thread. It segfaulted as soon as the path was genuinely exercised.

The store is now sized once from R, before the fit, from the subject count and
the maximum observations per subject. The hook only writes, and distinct
`(id, k)` land in distinct slots. The flat index is `id * stride + k`, and the
**same stride must be used on both sides** — it was once a `1024` literal in C
and in three places in R, which silently dropped any subject with more.

> **The reported score is on the TRANSFORMED scale.** Transform-both-sides is
> applied in rxode2's `rx_pred_` statement, so by the time the hook sees `f`, it
> is already transformed. The augmented solve's sensitivities are on the natural
> scale. Chaining them without the transformation's own derivative leaves the
> gradient short by that factor at every observation.

`R/nnEndpoint.R` supplies that derivative for every level in rxode2's table.
Note that a combined level is a *composition*: `"logit + yeoJohnson"` is
`yeoJohnson(logit(y))`, so the outer derivative is evaluated at `logit(y)`, not
at `y`. Multiplying two derivatives both taken at `y` is the plausible-looking
wrong answer.

### Which score is used, and why

| endpoint | score |
|---|---|
| untransformed, `add()`/`prop()` | closed-form Gaussian |
| anything else | exact, captured from the hook |

This is a decision, not a stopgap, and it is worth not re-litigating. The exact
score is evaluated at the *inner fit's* prediction while the sensitivities it
multiplies come from the *augmented* solve, and those two differ by ~4.5e-04 at
identical EBEs. The closed form takes both from the same solve, so where it
applies it is the more self-consistent of the two. The exact score earns its
place on the endpoints the closed form cannot express.

When capture fails, falling back to the closed form is only permitted where the
closed form is genuinely that endpoint's score — `.nnCanUseClosedForm()`.
Otherwise it errors, because a silent fallback on a transformed endpoint trains
on the wrong gradient.

## Model augmentation

`nnAugmentModel()` builds the forward-sensitivity model: for each weight `w_j`,
a variational state per model state,

```
d/dt(s_ij) = sum_k F_X[i,k] s_kj + (dR_i/dg)(dg/dw_j)
```

where `F_X` is the model Jacobian (rxode2 forms it, chaining the registered
`nn<K>_d*` input derivatives), `rx_drdg_<state>_` is emitted here, and `dg/dw_j`
comes from the plain-C `nnWeightGrad` at solve time. The prediction's
sensitivity is chained through the states as
`rx_predsw_j = sum_s d(pred)/d(s) * rx_sw_s`.

This costs **one extra ODE state per weight per model state**, which is the
current ceiling on network size.

Augmentation does not perturb the base trajectory — measured at ~5e-09 relative,
which matters because the closed-form score is computed from the augmented
solve.

```{r, eval = FALSE}
## the augmented text for a two-state model, for inspection
cat(nlmixr2nn:::nnAugmentModel(
  "g = nn1(0, centr)\nd/dt(centr) = -g*centr\nd/dt(peri) = g*centr - k*peri",
  H = 1L))
```

## The training backend

The modules live in C++ so compiled solve code reaches them without an R
callback. The backend is the *optimizer and the weight container*, not the
autodiff: the in-solve `dg/dw` uses the plain-C `nnWeightGrad`, which is
thread-safe and keeps the backend out of the ODE hot loop. `nnTorchSetGrad()` +
`nnTorchStep()` inject an externally computed gradient and take an Adam/SGD
step.

Because that is all a fit asks of it, the backend is interchangeable, and
`./configure` picks one at install time:

| | `src/nnTorch.cpp` | `src/nnBuiltin.cpp` |
|---|---|---|
| selected when | the 'torch' package has its libtorch binaries | otherwise, and by default on Windows |
| optimizer | libtorch's Adam/SGD | the same Adam/SGD, written out |
| autograd | real | the analytic VJP, from `nnWeightGradWC` |

`_nlmixr2nn_nnBackend()` reports which one is compiled in; only one ever is.
The fallback is why the package installs where libtorch cannot be linked --
Windows, where torch's libtorch is MSVC-built and Rtools is MinGW, and CRAN,
which will not fetch a 2GB library at build time.

The analytic weight gradient is checked against torch autograd across every
activation and input dimension, which is what licenses using the C path. That
check needs a libtorch build to mean anything -- under the builtin backend both
sides of it are the same routine -- so `test-nn-wgrad.R` skips unless libtorch
is linked, and CI keeps one row that links it.

## Persistence

A fitted model carries its trained weights in `rxForcedPars()` and its shapes in
a sticky `nnMeta`, so it reloads in a fresh session with no external state. The
transient C registry is rebuilt by the ui-prep hook, resolving the offset by
name against the solve model.

A model that carries its own weights is **self-describing, and wins over the
loader**: in a fresh session the loader buffer is empty, so letting it override
would silently zero the network. During training that is inverted — the loop
owns the weights and injects them every round.

## Things that look safe and are not

A short list, each of which cost real time:

- **Unexporting the generated `nn<K>` family.** Compiled model code reaches them
  through `R_RegisterCCallable()`, so it looks internal. But rxode2's symengine
  renderer resolves a model function *by name over the search path* when it
  builds derivatives, so an unexported `nn1` breaks every Jacobian.
  `nnWg<K>` is genuinely internal because it only ever appears in generated text
  handed to C.
- **Testing with `env = new.env(parent = asNamespace())`.** Package internals are
  visible there and not to a user, which hid the above completely.
  `test-nn-userenv.R` runs in a child of `globalenv()` and in a `callr`
  subprocess for exactly this reason.
- **Reading `rxForcedPars` as "supplied".** It is applied when gpars are written,
  *after* parameter resolution has decided whether every parameter has a source.
  rxode2 gained an explicit change so a forced parameter counts as supplied.
- **A tryCatch around an unexported call.** `nlmixr2est::nlmSolveR` does not
  exist (it is internal); the error was swallowed by a `tryCatch` returning `NA`,
  so an entire code path silently never ran.
- **The two weight-gradient sites are on different objective scales.**
  `R/nnWeightStep.R` hands torch a gradient of $-\log L$; `R/nnPopFit.R`
  optimizes $-2\log L$. The weight penalty (below) is defined on the
  $-2\log L$ scale, so adding the same vector at both would make one `l2` mean
  two different things inside a single fit, with nothing in the output to say so.
  `.nnAddPen(g, w, pen, scale)` takes the scale explicitly and
  `test-nn-penalty.R` asserts the invariant
  `2 * .nnAddPen(g, w, pen, 1L) == .nnAddPen(2 * g, w, pen, 2L)`.
- **Plain `sum(W1^2)` is not L2 here.** Input scaling is folded into the
  first-layer weights and then discarded (`.nnRescaleW1`), so an input of typical
  magnitude 500 carries a `W1` column ~500x smaller at equal functional effect.
  A naive weight-decay term would penalize it 250000x less. The penalty is on the
  *effective* weight `s_k * W1[j,k]`, which is why the spec carries the input
  scales.
- **`nnWeightGrad` outside a solve returns zeros, not an error.** It reads the
  live solve's `par_ptr`. Anything evaluating a network outside a solve — the
  curvature penalty, `nnEval()` — must use the explicit-weight entry points
  `_nlmixr2nn_nnForwardW` / `_nlmixr2nn_nnWeightGradW`. Using the registry form
  would make the whole penalty vanish silently.

## The weight penalty

`R/nnPenalty.R`. Two terms, both opt-in (`nnControl(l2 = , smooth = )`, both `0` by default):

$$P = -2\log L + \lambda_{2}\sum W_{\text{eff}}^2 + \lambda_{s}\sum C^2$$

`C` is the second difference of the network along one input's marginal curve,
that input swept over its observed 10th-90th percentile with the others pinned at
their center. Sweeping each input's *own* range is what makes `smooth` scale-free
without extra normalization: $C = h^2 f''(x)$ and $h$ is proportional to the
input's spread, so `C` is curvature with respect to the standardized input.

Inputs with no resolvable range — a latent eta, whose EBEs move every round, or a
compound expression such as `nn(centr/Vc)` — are held fixed rather than swept.

The gradient reuses the analytic weight Jacobian:
$\partial C/\partial w = J(x_{i+1}) - 2J(x_i) + J(x_{i-1})$, so there is no new
autodiff and no new C code.

`nnControl(l2 = 0, smooth = 0)` is a strict no-op: the spec is `NULL`, the
gradient vector is returned untouched, and the fit is bit-identical to an
unregularized one.

### Lambda is a fraction, not an amount

Each term carries a normalizer, frozen at the first evaluation that knows an
objective (the population pre-fit's own objective, or the first inner fit in the
round loop):

$$k_W = \frac{|\text{objf}|}{\sum W_{\text{eff}}^2}, \qquad
  k_C = \frac{|\text{objf}|}{\sum C^2}$$

so that each term *starts* at exactly $\lambda$ times the objective. They are
frozen rather than recomputed, so $P$ stays a fixed objective a gradient can
descend rather than a moving target; until they are known the penalty is
inactive and contributes exactly zero.

The two terms are normalized **separately** because their raw scales differ by
orders of magnitude — measured at ~180× — so one shared normalizer would leave
the mix between them as arbitrary as the magnitude was.

This is not cosmetic. Under the earlier absolute penalty the E5.5 smoke matrix
broke at `l2 = 0.1` (the `saem`/`lnorm` cell's RMSE rose instead of falling)
while visible shrinkage on an over-fitting fit needed `l2 = 1` — the safe range
and the effective range did not overlap. After normalizing, the whole matrix
passes to `l2 = 0.3` and the objective-optimal value is `0.05`.

## Extending

- **An activation**: `.nnActCode` (R), `nnAct`/`nnActD`/`nnActD2` in
  `src/nnEval.c`, and `MLPImpl` in `src/nnTorch.cpp`. The three must agree;
  `test-nn-wgrad.R` and `test-nn-general.R` check that they do.
- **More inputs than 4**: regenerate `R/nnGen.R` with `tools/genNn.R` and raise
  `NN_KMAX`.
- **Another estimator**: teach `.nnEstSolveParams()` which model it solves, and
  `.nnInterleaveKnob()`/`.nnNonResuming` whether its partial fit resumes.
- **Depth**: `.nnInitDraw()` and the weight layout already take `H` as a vector;
  the compiled evaluators do not. `nn()` refuses a vector `nHidden` explicitly
  so that boundary is visible.

## Where to look when something is wrong

The tests are organised around the invariants rather than the functions:

| file | what it pins |
|---|---|
| `test-nn-consistency.R` | the network evaluated == the network the weights describe |
| `test-nn-reproducible.R` | a fit is reproducible under a fixed seed |
| `test-nn-userenv.R` | it works with only the package attached |
| `test-nn-endpoint.R` | the transformation derivative, vs rxode2's own transform |
| `test-nn-llgrad.R` | the assembled gradient, vs a finite difference of the likelihood |
| `test-nn-cotangent-agree.R` | the two score sources, compared at the gradient |
| `test-nn-zeroboiler.R` | no helper call is ever required |
