---
title: "Neural-network ODEs with nlmixr2nn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Neural-network ODEs with nlmixr2nn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
.hasRx    <- requireNamespace("rxode2", quietly = TRUE)
.hasTorch <- .hasRx &&
  tryCatch(isTRUE(.Call("_nlmixr2nn_nnTorchAvailable")), error = function(e) FALSE)
```

## What nlmixr2nn does

`nlmixr2nn` lets you drop a neural network into an ODE right-hand side for
`rxode2`/`nlmixr2`. It extends the idea of `pmxNODE`, but instead of expanding
every weight into explicit model algebra (which does not scale), the network is
an **opaque, compiled function**: the model text stays O(1) in the network size,
the value and its analytic input derivatives are computed in fast thread-safe C,
and a C++ `libtorch` module owns and trains the weights.

The network is written with the lower-case `nn()` function (upper-case `NN()` is
`pmxNODE`):

```{r, eval = .hasRx}
library(nlmixr2nn)

mod <- function() {
  ini({
    lka <- 0.5
    lVc <- 1
  })
  model({
    ka <- exp(lka)
    Vc <- exp(lVc)
    d/dt(depot)   <- -ka * depot
    ## a 6-unit softplus network of the central amount enters the elimination
    d/dt(central) <-  ka * depot - nn(central, n_hidden = 6, act = "softplus")
    cp <- central / Vc
  })
}
ui <- rxode2::rxode2(mod)
```

At parse time `nn(central, ...)` is replaced by a single compiled call and the
network weights are added as randomly-initialized parameters:

```{r, eval = .hasRx}
cat(paste(vapply(ui$lstExpr, deparse1, character(1)), collapse = "\n"))
```

## Solving with a torch module

Create a `libtorch` module for each network in the model, register the weight
layout, and solve. `nnTorchModel()` builds the modules; `nnUpdate()` resolves
where each network's weights live in the solve parameter vector and syncs the
module weights into the loader buffer that the rxode2 hook injects each solve.

```{r, eval = .hasTorch}
nnTorchModel(ui, seed = 1)   # C++ torch module for network 0
nnUpdate(ui)                 # register layout + sync weights

ev <- rxode2::et(amt = 100, cmt = "depot") |> rxode2::et(0, 24, by = 4)
head(rxode2::rxSolve(ui, ev))
```

## Training the network

The weights are trained by torch, not by nlmixr2's optimizer. The gradient
bridge turns a cotangent on the network output into `d(loss)/d(weights)` via
torch autograd and steps a torch optimizer. `nnTorchTrain()` demonstrates this
with an explicit L2 target (in a full fit the cotangent comes from the
likelihood instead):

```{r, eval = .hasTorch}
id <- 7L
nnTorchInit(id, K = 1L, H = 12L, act = "gelu", seed = 3)
X <- matrix(seq(-2, 2, length.out = 41), ncol = 1)
loss <- nnTorchTrain(id, X, target = sin(X), steps = 400L, lr = 0.05)
c(first = loss[1], last = tail(loss, 1))
```

## Population fitting: the transparent workflow

You fit a mixed-effects model whose ODE contains an `nn()` term the *same way you
fit any other* `nlmixr2` model -- with a **standard estimator** (`"focei"`,
`"saem"`, ...). Loading `nlmixr2nn` registers an estimation interceptor that
detects the `nn()` term and trains the embedded network transparently, using the
estimator you chose as the inner engine. The only extra argument is `nn =
nnControl(...)`, the network training *schedule* (it does **not** wrap the inner
control):

```{r, eval = FALSE}
nlmixr2(model, data, "focei", foceiControl(...), nn = nnControl(...))
```

`nnControl(mode=)` picks how the network and the population parameters are
optimized:

- **`mode = "joint"`** (default) -- co-optimizes the population parameters and the
  network weights in one interleaved loop: each round takes a few *partial*,
  warm-started inner iterations and a torch weight step, so the parameters and
  weights descend *together*. This is the DeepPumas-style approach. True
  interleaving needs an inner estimator whose partial fit *resumes* from the
  previous round's estimates: the FOCEi family (`maxOuterIterations`) and the
  variational methods `advi`/`vae` (`iters`). SAEM/fSAEM (whose stochastic-
  approximation chain restarts each call) and `imp`/`qrpem` degrade to the
  iterative loop, which is just as good for them.
- **`mode = "iter"`** -- runs a *full* inner fit each round with the weights held
  fixed, then updates the weights: a simpler solve/update/solve scheme. Useful
  when you want each round to be an ordinary, fully converged fit.

Both stop when the weights (and, for `"joint"`, the objective) stop changing
between rounds -- `nnControl(tol=)` -- or at the maximum `rounds`. The returned
fit reports `$nnConverged`, `$nnRounds`, and `$nnParHist` (the per-round trace).
Covariance and output tables are added once, post-hoc, at the trained weights (the
intermediate rounds skip them).

**Universal differential equation (UDE) correction** -- keep the mechanistic
structure you trust and let a small network absorb the part you do not:

```{r, eval = FALSE}
ude <- function() {
  ini({ tcl <- 1; tv <- 3.5; add.sd <- 0.5; eta.nn ~ 0.1 })
  model({
    cl <- exp(tcl); v <- exp(tv)
    d/dt(central) <- -(cl / v) * central + nn(central, eta.nn)  # learned correction
    cp <- central / v
    cp ~ add(add.sd)
  })
}
```

**Covariate model discovery** -- let the network learn how covariates drive an
individual parameter, with a latent random effect for the residual variation:

```{r, eval = FALSE}
covmod <- function() {
  ini({ tCL <- 1; tV <- 3.5; add.sd <- 0.5; eta.nn ~ 0.1 })
  model({
    CL <- exp(tCL + nn(WT, EGFR, eta.nn))                       # learned covariate map
    V  <- exp(tV)
    d/dt(central) <- -(CL / V) * central
    cp <- central / V
    cp ~ add(add.sd)
  })
}
```

In both cases the inter-individual variability enters as a *latent input* to the
population network (`eta.nn` above): its inner sensitivity is the network's
analytic input derivative, so it is identifiable and needs no finite differences.

You fit with an ordinary estimator (here FOCEi) and its ordinary control; the
neural-network schedule (`mode`, `wSteps`, `lr`, `tol`, ...) rides alongside in
`nn = nnControl(...)`:

```{r, eval = .hasTorch}
nnClearMeta()
mm <- function() {
  ini({ add.sd <- 0.3; eta.nn ~ 0.2 })
  model({
    g <- nn(centr, eta.nn, n_hidden = 3L, act = "tanh")
    d/dt(centr) <- -(1.0 / (1.0 + exp(-g))) * centr
    centr ~ add(add.sd)
  })
}
set.seed(1)
d <- do.call(rbind, lapply(1:6, function(id) {
  data.frame(id = id, time = c(0, 1, 2, 4, 6), evid = c(1, 0, 0, 0, 0), cmt = 1,
             amt = c(10, 0, 0, 0, 0),
             dv = c(NA, 8, 6, 4, 2) + rnorm(5, 0, 0.1))
}))
fit <- nlmixr2est::nlmixr2(mm, nnCovData(d), "focei",
                           nlmixr2est::foceiControl(print = 0L, maxInnerIterations = 10L),
                           nn = nnControl(mode = "joint", rounds = 10L, wSteps = 2L,
                                          lr = 0.03, tol = 1e-2, seed = 5))
```

The result is a standard `nlmixr2` fit: `fit$eta` holds the per-subject latent
EBEs, `fit$nnParHist` the round-by-round training trace, and the trained weights
are baked into the model as `rxForcedPars()` -- so `predict(fit)` and simulation
reproduce them with no external torch/loader state (the fit is self-contained and
portable, like a `nonmem2rx` model).

The inner estimator is whatever standard `est` you pass: `est = "saem"` or `est =
"imp"` run the network training with SAEM or importance-sampling inner fits (the
network weights reach the inner fit through the data, so no estimator-specific
plumbing is needed). If the fitted model already carries trained weights (as
`rxForcedPars`, e.g. a previous `nlmixr2nn` fit), those are used as the warm-start
rather than a fresh random initialization.

**Warm starts.** Random weight initializations can be fragile, so a population
weight pre-fit -- the "nlm bridge" -- runs **by default** before the loop (unless
the model already carries trained weights, e.g. a fit passed back in, in which case
those are the warm start and the pre-fit is skipped). It fits the weights at `eta =
0` with a fixed-effects optimizer: the weights sit in the vector the optimizer
moves, giving a robust population weight vector before the random effects enter. As
*nlm is simply an optimizer*, `nnControl(warmStart = ...)` picks any nlm-family
optimizer -- `"lbfgsb3c"` (default), `"nlminb"`, `"nlm"`, `"optim"`, `"n1qn1"`
(gradient-based, using the analytic sensitivity gradient) or `"bobyqa"`,
`"newuoa"`, `"uobyqa"` (derivative-free) -- and none uses a random effect. It is
kept deliberately light (`warmPopIters = 3`): a few iterations escape a bad
initialization while leaving the per-subject variation for the random effect, since
a long pre-fit lets the population network absorb IIV that should go to the eta.
Set `warmStart = "none"` to disable it, or raise `warmPopIters` for a stronger
fixed-effects start (e.g. a population/UDE model with no random effect).
`nnControl(warmSteps = n)` additionally takes `n` naive-pooled torch steps.

## Inputs and activations

The compiled family supports input dimensions `K = 1..4` -- pass several states
positionally, e.g. `nn(central, t)` -- and the activations `"relu"`,
`"softplus"`, `"tanh"`, `"gelu"`, and `"silu"`. Value, input gradient, and input
Hessian are analytic in every case.

## Saving and restoring a trained network

A trained module serializes to a binary and restores for later simulation:

```{r, eval = .hasTorch}
f <- tempfile(fileext = ".pt")
nnTorchSave(id, f)
nnTorchInit(id, K = 1L, H = 12L, act = "gelu")   # reinitialize (random)
nnTorchLoad(id, f)                               # restore trained weights
unlink(f)
```
```{r, include = FALSE}
if (.hasTorch) try(nnTorchFree(id), silent = TRUE)
```
