---
title: "Neural networks in nlmixr2 models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Neural networks in nlmixr2 models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
## NB: the package must be loaded before probing the backend -- the symbol lives
## in its DLL, so probing first silently reports "no torch" and skips every
## fitting chunk, leaving a vignette that builds in seconds and demonstrates
## nothing.
.hasRx    <- requireNamespace("rxode2", quietly = TRUE)
.hasNn    <- .hasRx && requireNamespace("nlmixr2nn", quietly = TRUE)
.hasEst   <- .hasNn && requireNamespace("nlmixr2est", quietly = TRUE)
.hasTorch <- .hasEst &&
  tryCatch(isTRUE(.Call("_nlmixr2nn_nnTorchAvailable", PACKAGE = "nlmixr2nn")),
           error = function(e) FALSE)
```

## What `nn()` is

`nn()` puts a neural network inside a model. You write it where you would write
any other term, and everything else is an ordinary `rxode2`/`nlmixr2` workflow:
the model solves, `nlmixr2()` fits it, `predict()` works, `saveRDS()` works.

```{r, eval = .hasRx}
library(nlmixr2nn)

mod <- function() {
  ini({
    lka <- 0.5
    lVc <- 1
  })
  model({
    ka <- exp(lka)
    Vc <- exp(lVc)
    d/dt(depot)   <- -ka * depot
    ## a 6-unit network of the central amount enters the elimination
    d/dt(central) <-  ka * depot - nn(central, n_hidden = 6)
    cp <- central / Vc
  })
}

set.seed(42)
ui <- rxode2::rxode2(mod)
```

The network is **opaque**: it becomes a single compiled call, so the model text
stays the same size no matter how big the network is. Its value and its exact
input derivatives are computed in fast thread-safe C, and those derivatives are
what make a random effect passed *into* a network identifiable.

```{r, eval = .hasRx}
cat(paste(ui$lstChr, collapse = "\n"))
```

The weights were drawn when the model was parsed, using R's random number
generator — so `set.seed()` reproduces a network exactly, in this session or any
other, and the model is ready to solve immediately.

```{r, eval = .hasRx}
head(nnWeights(ui), 4)
```

Everything below is a single `nlmixr2()` call. There is no setup step.

## Simulating a model you have not fitted

Because the model carries its own weights, it solves straight away. This is the
quickest way to check that a model does what you think before fitting anything.

```{r, eval = .hasRx}
ev <- rxode2::et(amt = 100, cmt = "depot") |> rxode2::et(0, 24, by = 4)
head(rxode2::rxSolve(ui, ev), 4)
```

## Learning a term you do not know (UDE)

Keep the mechanism you trust and let a small network absorb the part you do not.
Here the truth is a saturating (Michaelis-Menten) elimination that varies between
subjects; the model knows only that elimination is *some* bounded function of the
amount, and a latent random effect `eta.nn` enters the network as an input.

```{r, eval = .hasTorch}
set.seed(1)
truth <- rxode2::rxode2("d/dt(centr) = -(2*exp(eV))*centr/(3+centr)")
etaTrue <- rnorm(8, 0, sqrt(0.15))
d <- do.call(rbind, lapply(1:8, function(id) {
  s <- rxode2::rxSolve(truth,
         data.frame(id = id, time = c(0, .5, 1, 2, 4, 6, 8, 10),
                    evid = c(1, rep(0, 7)), cmt = 1, amt = c(10, rep(0, 7))),
         params = c(eV = etaTrue[id]), returnType = "data.frame")
  s <- s[s$time > 0, ]
  data.frame(id = id, time = c(0, s$time), evid = c(1, rep(0, nrow(s))), cmt = 1,
             amt = c(10, rep(0, nrow(s))), dv = c(NA, s$centr + rnorm(nrow(s), 0, .1)))
}))

ude <- function() {
  ini({ add.sd <- 0.3; eta.nn ~ 0.2 })
  model({
    g <- nn(centr, eta.nn, n_hidden = 3, act = "tanh")
    d/dt(centr) <- -(1 / (1 + exp(-g))) * centr     # a bounded learned rate
    centr ~ add(add.sd)
  })
}

fit <- nlmixr2est::nlmixr2(ude, d, "focei")
```

That is the whole call: no `nn=` argument, and the data was not touched. The
result is an ordinary `nlmixr2` fit, so the random effect recovers the
between-subject variation the way any eta would:

```{r, eval = .hasTorch}
cor(fit$eta$eta.nn[order(fit$eta$ID)], etaTrue)
```

To see what the network *learned*, evaluate it directly:

```{r, eval = .hasTorch}
e <- nnEval(fit, centr = seq(0.5, 10, length.out = 25), eta.nn = 0)
head(e, 3)
```

`plot(e)` draws that function, and `plot(e, true = f)` overlays a reference
curve when you have one.

## Letting the network find a covariate relationship

The same idea with covariates as the inputs: the network learns how they drive a
parameter, and the latent eta carries what they do not explain.

```{r, eval = FALSE}
covmod <- function() {
  ini({ tCL <- 1; tV <- 3.5; add.sd <- 0.5; eta.nn ~ 0.1 })
  model({
    CL <- exp(tCL + nn(WT, EGFR, eta.nn))        # learned covariate map
    V  <- exp(tV)
    d/dt(central) <- -(CL / V) * central
    cp <- central / V
    cp ~ add(add.sd)
  })
}
fit <- nlmixr2est::nlmixr2(covmod, data, "focei")
```

## A model with no between-subject variability

Systems-pharmacology models are often fitted with no random effects at all: one
mechanistic system, with a network standing in for an unknown term. Fit those
with a population optimizer — `"bobyqa"`, `"nlminb"`, `"lbfgsb3c"` and friends —
and there is no eta to estimate.

```{r, eval = .hasTorch}
set.seed(1)
qspTruth <- rxode2::rxode2("d/dt(centr) = -(2)*centr/(3+centr)")
qd <- do.call(rbind, lapply(1:6, function(id) {
  s <- rxode2::rxSolve(qspTruth,
         data.frame(id = id, time = c(0, .5, 1, 2, 4, 6, 8, 10),
                    evid = c(1, rep(0, 7)), cmt = 1, amt = c(10, rep(0, 7))),
         returnType = "data.frame")
  s <- s[s$time > 0, ]
  data.frame(id = id, time = c(0, s$time), evid = c(1, rep(0, nrow(s))), cmt = 1,
             amt = c(10, rep(0, nrow(s))), dv = c(NA, s$centr + rnorm(nrow(s), 0, .1)))
}))

qsp <- function() {
  ini({ add.sd <- 0.3 })                     # residual error only -- no eta
  model({
    g <- nn(centr, n_hidden = 3, act = "tanh")
    d/dt(centr) <- -(1 / (1 + exp(-g))) * centr
    centr ~ add(add.sd)
  })
}

qspFit <- nlmixr2est::nlmixr2(qsp, qd, "bobyqa")
```

The fitted model reproduces the dynamics it was never told about:

```{r, eval = .hasTorch}
nd <- data.frame(time = c(0, .5, 1, 2, 4, 6, 8, 10), evid = c(1, rep(0, 7)),
                 cmt = 1, amt = c(10, rep(0, 7)))
a <- rxode2::rxSolve(qspFit$finalUi, nd, returnType = "data.frame")
b <- rxode2::rxSolve(qspTruth, nd, returnType = "data.frame")
data.frame(time = a$time, fitted = round(a$centr, 2), true = round(b$centr, 2))
```

Passing a population-only optimizer for a model that *does* have a random effect
is an error rather than a silent compromise — the weights and the Omega would
otherwise come from two different objectives.

## Saving, reloading and refitting

The trained weights are carried on the fitted model, so a saved fit is
self-contained: it reloads and re-solves in a fresh session with no external
state. Feeding a fit back in resumes from its trained weights.

```{r, eval = .hasTorch}
f <- tempfile(fileext = ".rds")
saveRDS(qspFit, f)
reloaded <- readRDS(f)
identical(round(rxode2::rxSolve(reloaded$finalUi, nd, returnType = "data.frame")$centr, 8),
          round(a$centr, 8))
unlink(f)
```

```{r, eval = FALSE}
refit <- nlmixr2est::nlmixr2(qspFit, moreData, "bobyqa")   # warm start
```

## When you want to steer the training

The training schedule is inferred from the model, the data and the estimator you
chose: whether the estimator can resume from a partial fit, whether there is a
random effect, and what the residual model is. `nnControl()` overrides any part
of it and leaves the rest inferred.

```{r, eval = FALSE}
nlmixr2est::nlmixr2(ude, d, "focei", nn = nnControl(rounds = 400, lr = 0.01))
```

| setting | inferred as | override when |
|---|---|---|
| `mode` | `"joint"` when the estimator resumes from a partial fit, else `"iter"` | you want fully converged inner fits each round |
| `rounds` | 200 joint, 30 iterative, 60 population | training was still improving when it stopped |
| `cotangent` | the Gaussian closed form for an untransformed `add()`/`prop()` endpoint, the distribution's own score for a count endpoint, the captured exact score otherwise | rarely — a source the endpoint cannot support is an error, not a silent fallback |
| `lr`, `optimizer` | 0.03 (0.01 on a refit), Adam | the objective oscillates or moves too slowly |
| `warmStart` | a population pre-fit, skipped when the model already carries trained weights | you want a longer or shorter pre-fit |
| `l2`, `smooth` | a light penalty on the weights and on curvature, applied by default | the network still looks over-fitted, or you want the unpenalized fit back |

### Regularization is on by default

A network with more capacity than the data supports will invent structure — a
covariate curve with sharp wiggles in it, or per-subject variation that should
have gone to the random effect. It does this silently: the fit converges, the
objective looks fine, and only the learned shape gives it away.

So two penalties are applied by default. `l2` shrinks the weights; `smooth`
charges for curvature specifically, penalizing the *second* derivative of the
network along each input, which leaves a monotone slope free and pays only for
the wiggles. Biases are never penalized, so the network can still move its output
level for nothing.

Both shape the optimization only. The reported `objf` — and so AIC and BIC — stays
the unpenalized -2 log-likelihood, which means a regularized network model is
still directly comparable to one with an analytic covariate relationship.

To turn regularization off completely:

```{r, eval = FALSE}
nlmixr2est::nlmixr2(ude, d, "focei", nn = nnControl(l2 = 0, smooth = 0))
```

That reproduces an unpenalized fit exactly — zero is a strict no-op, not a small
number.

The defaults are set conservatively: low enough not to disturb any supported
estimator and endpoint combination, which is lower than the value that does the
most good on a model that really is over-fitting. So if a fitted curve still
looks wigglier than the data can justify, raise `l2` — around `1` is where
shrinkage becomes substantial — and check the result with `nnEval()` and
`plot()`:

```{r, eval = FALSE}
fit <- nlmixr2est::nlmixr2(ude, d, "focei", nn = nnControl(l2 = 1))
plot(nnEval(fit, centr = seq(0, 10, length.out = 101)))
```

## What is supported

- **Inputs**: 1 to 4, given positionally — states, covariates, or a latent eta.
- **Activations**: `"softplus"` (default), `"tanh"`, `"relu"`, `"gelu"`, `"silu"`.
  The default is smooth with a nonzero second derivative, which matters because
  the model Jacobian and the FOCEi sensitivities are chained through it.
- **Hidden layers**: one. `n_hidden` is a single width.
- **Endpoints**: one. Any normal-distribution residual model — `add()`, `prop()`,
  both, or a transformation such as `lnorm()` — and the count endpoints `pois()`
  and `binom()`.

  A count endpoint differs in one way worth knowing. For `y ~ pois(lam)` there is
  no "prediction" to differentiate: the model's prediction *is* the log-density,
  so the network's effect is chained through the distribution's own parameter
  (`lam`) instead, and the score is that distribution's derivative. Two
  consequences: `nnControl(cotangent = "exact")` does not apply and is refused,
  and **the parameter must be positive by construction** — write
  `lam <- exp(nn(...))` rather than letting a rate wander negative, which aborts
  the likelihood evaluation. For `binom(n, p)` the network drives `p`; the size
  `n` must reach the solve, so bind it to a data column (`n <- ntrials`) rather
  than writing a bare constant, which the model compiler folds away.

  Other distributions (`nbinom()`, ordinal, and the rest) are refused with a
  message naming the reason, not fitted wrongly.
- **Networks per model**: several; they are trained together, and `nnEval(net=)`
  selects between them.
- **Estimators**: any standard one. `"focei"` and the variational methods
  interleave network training with the population fit; SAEM and the
  importance-sampling methods alternate instead.

Network initialization is scaled to each input's magnitude, measured from the
data, so a network reading a concentration of 0.02 and one reading an amount of
500 both start in a usable range.
