---
title: "Neural ODEs and nlmixr2nn: what carries over"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Neural ODEs and nlmixr2nn: what carries over}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`nn()` puts a neural network inside an ODE, which is what the neural-ODE (NODE)
literature is about. That literature is written for a different problem, though,
and the differences decide which of its results apply here and which do not.
This vignette is for readers who know that literature and want to know what
carried over.

The comparison throughout is against Worsham and Kalita (2025) and its
reference implementation, which is the most complete practical survey of the
area: solvers, differentiation modes, augmentation and regularization, each with
measured experiments.

## The problem is not the same one

In the NODE literature the network **is** the dynamics: `dx/dt = f(x; w)`, with
`f` a 32x32 multilayer perceptron, fitted to trajectories that are dense, cheap
and usually noise-free. A single trajectory is one training example and the loss
is often evaluated only at its endpoint.

Here the network is a **term inside** a model that is mostly known:

```{r, eval = FALSE}
d/dt(centr) <- -(1 / (1 + exp(-nn(centr, eta.nn)))) * centr
```

That is a universal differential equation (UDE): the mechanism you trust stays,
and a small network absorbs the part you do not. Four differences follow, and
they drive everything below.

* **Data is sparse.** Eight observations per subject, tens of subjects, with
  real residual error. A 32x32 network has more parameters than the study has
  observations.
* **Every observation is an output.** The loss is not a terminal condition; it
  accumulates at every measured time.
* **Individual gradients matter.** The population fit needs per-subject
  quantities, not just a summed gradient.
* **The likelihood is the objective.** Not mean squared error -- a residual
  model, random effects and an objective function value that has to stay
  comparable to a model fitted without a network.

## Sensitivities: forward, not adjoint

The adjoint method is the headline result of Chen et al. (2018): solve one
augmented system backwards in time and get the gradient with respect to every
parameter at a cost independent of how many parameters there are.

`nlmixr2nn` does not use it. The weight gradient comes from **forward
sensitivities** -- `nnAugment.R` adds one variational state per (state x weight)
to the solve, so `d(state)/d(w)` is integrated alongside the states themselves.

That is a deliberate choice, for three reasons in decreasing order of weight:

1. **The adjoint's win condition does not hold.** It pays when outputs are far
   fewer than parameters -- ideally one scalar loss. With an observation at every
   measured time, a backward sweep must be re-seeded at each one, and it returns
   only the reduced sum-gradient. One forward pass returns every observation's
   sensitivity to every weight.
2. **Forward sensitivities are already being computed.** FOCEi needs
   `d(pred)/d(eta)`, which is a forward variational solve using the same model
   Jacobian `F_X`. Adding weight columns reuses it. An adjoint would be a
   second, parallel, backward machinery sharing none of that work.
3. **The measurements do not favour it.** In Worsham and Kalita's own pendulum
   experiments -- on a terminal-only loss, the setup most favourable to the
   adjoint -- it was slightly slower than direct differentiation and no more
   accurate.

This is not a statement that adjoints are unavailable in this ecosystem.
`rxode2` implements exact discrete adjoint sensitivities, including stiff
methods and steady-state dosing, aimed at the **THETA** gradient in `nlm`/`focei`
-- where the objective really is one scalar and the parameters really are many.
That is the case the adjoint was built for. A network's weight gradient is not.

### What that costs, and what follows from it

The forward path's cost is `O(nStates x nWeights)`, and since it is not going
away, it is a design constraint rather than a temporary limit. Two consequences
are worth stating plainly, because they run against the literature's defaults.

**Prefer width to depth.** A single hidden layer of width `H` with `K` inputs is
`H*K + 2H + 1` weights; a second layer adds a quadratic `H1*H2` block:

| network (K = 2) | weights, and so variational states per model state |
|---|---|
| `n_hidden = 5` (the default) | 21 |
| `n_hidden = 32`, one layer | 129 |
| `n_hidden = c(16, 16)` | 337 |

Roughly the same capacity, 2.6x the sensitivity cost for the two-layer version.
`nn()` accepts a single hidden layer for this reason, and rejects a vector
`n_hidden` at the user boundary rather than deeper down, so depth remains an
additive change if the trade ever becomes worth making.

**Prefer small networks generally.** With eight observations per subject, the
32x32 default of the NODE literature is not a network this data can support --
the constraint and the statistics point the same way for once.

## Regularization

Three penalties are available, and they act in different spaces. All three shape
the optimization only: the reported `objf`, and so AIC and BIC, stays the
unpenalized -2 log-likelihood, so a regularized network model is still directly
comparable to one with an analytic covariate relationship.

| penalty | acts on | where it is measured | default |
|---|---|---|---|
| `l2` | weight magnitude | the weights themselves | 0.01 |
| `smooth` | curvature | a synthetic partial-dependence grid | 1 |
| `kinetic` | how hard the network pushes | the realized trajectory | 0 (off) |

`l2` and `smooth` are described in the main vignette. `kinetic` comes from this
literature and is the one worth explaining here.

### The kinetic term

`kinetic` adds `kinetic * mean(f^2)`, where `f` is the network's output at the
**jointly realized input rows** of a trial solve -- the trajectory the model
actually produces over the real dosing and time grid. It is the kinetic-energy
or optimal-transport regularizer of Finlay et al. (2020): it asks the learned
dynamics to reach the same data by pushing less hard.

```{r, eval = FALSE}
nlmixr2est::nlmixr2(ude, d, "focei", nn = nnControl(kinetic = 0.1))
```

The difference from `smooth` is where the question is asked. `smooth` sweeps one
input across its range with the others pinned at their center, and charges for
second differences: it asks whether the learned function is wiggly *in the
abstract*, at input combinations that may never occur. `kinetic` asks how hard
the network is pushing *where the solution really goes*, which makes it a
statement about the dynamics the solver has to integrate. Inputs the trial solve
does not carry -- an eta, or a compound expression such as `nn(central/Vc)` --
are held at their center rather than paired with rows they never occurred with;
fabricating input combinations is exactly what this penalty exists to avoid.

Two details are deliberate. The penalty is a **mean**, where `smooth` is a sum:
the curvature grid is a fixed 11 points, so summing over it is
dataset-independent, but the kinetic point set is however many rows the solve
produced, and summing there would make one lambda mean something different for
every study. And it is `f^2` rather than the plain norm the reference
implementation uses, because the norm is not differentiable at zero and a UDE
network initialized as a small perturbation of the mechanistic model
(`init = "ude"`, the default) starts very close to exactly there.

It is **off by default**, unlike the other two, and the asymmetry is the point:
`l2` and `smooth` steer only the optimizer, while `kinetic` changes the dynamics
being solved. A default that quietly flattens a learned elimination rate would
be a default that quietly changes the model.

Two things recommend turning it on. In Worsham and Kalita's comparison it was
the single largest improvement of any lever they measured -- roughly halving the
error -- and it made training *faster*, because the learned dynamics became
easier to integrate. That second effect is amplified here: the augmented solve
carries a variational state per state-weight pair, so every solver step a
smoother right-hand side saves is multiplied by the whole sensitivity block.
Reach for it when the augmented solve is slow, or when the learned term looks
stiffer than the data justifies.

## Augmentation, and a limit worth knowing

The clearest theoretical result in this literature is that NODE trajectories
cannot cross. A flow in `n` dimensions is a homeomorphism at each time, so some
maps are simply not representable, and no amount of training or capacity fixes
it. Dupont et al. (2019) demonstrate this with the XOR and nested-spheres
problems and solve it by **augmenting** the state with extra dimensions
initialized at zero; Norcliffe et al. (2020) reach the same place by giving the
state a learned velocity, a second-order NODE.

This applies directly here. A one-state UDE

```{r, eval = FALSE}
d/dt(centr) <- -f(nn(centr))
```

is exactly the one-dimensional non-crossing case: the learned flow is
constrained in a way that has nothing to do with how well it is fitted.

`nlmixr2nn` has no first-class `aug =` argument yet, and the reason is worth
recording because it is a property of where `nn()` runs rather than of the
construction. A `nn()` term expands during parsing and can only emit code around
the line it sits on, so an augmented `d/dt()` lands next to the `nn()` call
rather than after every compartment the user declared. rxode2 numbers
compartments by first `d/dt()` appearance, so the augmentation *takes* a
compartment number, and `cmt = 1` in the data then doses the latent instead of
the model's first compartment. Measured on a UDE fit: the dose landed in the
augmented state, `centr` stayed at 0 for the entire profile, and the fit looked
like a training failure when nothing was ever being trained. A UDF cannot see
the model's states to prevent it (`rxUdfUiMv()` is `NULL` while parsing), so the
fix is to append the augmented compartments at model-assembly time, where the
whole model is visible.

Two other findings from that work are settled and will carry over:

* **The augmented state must relax, not integrate.** The textbook drift
  `d/dt(a) = f(x, a)` is self-exciting here -- `a` re-enters its own derivative
  through the network, and an unbounded activation makes that exponential. The
  latent reached 1e154 by `t = 5` and the solver gave up. `d/dt(a) = f(x, a) - a`
  has a stable fixed point at `f`, still adds the dimension the non-crossing
  argument needs, and costs a fixed memory timescale of one model time unit.
* **The sensitivity machinery already handles it.** The forward variational
  block for a two-network, two-state augmented model matches finite differences
  of the likelihood to ~1e-7 relative, for both networks
  (`test-nn-augsens.R`). Whatever else augmentation needs, the gradient is not
  it.

Meanwhile, two things reduce the sting, and neither is quite the same fix:

* **Extra inputs.** The recommended pattern passes a latent random effect into
  the network (`nn(centr, eta.nn)`), and covariates can be passed the same way.
  That augments the *input* space per subject, which is enough to separate
  trajectories that would otherwise coincide -- but it does not add dynamic
  dimensions, so it is not the ANODE construction.
* **Extra compartments.** Nothing stops you writing the augmented states
  yourself as ordinary compartments with `nn()` terms in their right-hand sides.
  This is the real construction, at the cost of writing it out -- and, written
  by hand, your compartments keep their numbers:

  ```{r, eval = FALSE}
  model({
    d/dt(a1)    <- nn(centr, a1) - a1     # the augmented dimension
    g           <- nn(centr, a1)
    d/dt(centr) <- -(1 / (1 + exp(-g))) * centr
  })
  ```

  Declare it after the compartments your data doses into, and give the relaxing
  `- a1` term its due: without it the latent runs away.

## What is deliberately absent

Some of the toolkit does not belong in a pharmacometric model, and that is worth
saying explicitly so its absence does not read as an omission.

* **Dropout.** It would make the ODE right-hand side stochastic, breaking both
  the deterministic solve and the analytic Jacobian `rxode2` needs. (The
  reference implementation offers it in one model class and lists it as an open
  question in its own notebooks.)
* **Learned nonlinear encoders and decoders.** Massaroli et al. (2020) warn that
  a nonlinear encoding before a NODE renders the NODE itself largely
  unnecessary, and identifiability here argues the same way. `nn()` does apply a
  fixed, data-derived diagonal input scaling -- a non-learned linear encoder --
  because a network fed raw PK-scale inputs starts saturated and neither trains
  nor yields an identifiable latent eta.
* **A solver zoo, and SDE integrators.** `rxode2`'s solvers are the ones used,
  and they are already stiff-capable and adaptive.

## References

Chen RTQ, Rubanova Y, Bettencourt J, Duvenaud D (2018). "Neural ordinary
differential equations." *Advances in Neural Information Processing Systems*
31.

Dupont E, Doucet A, Teh YW (2019). "Augmented neural ODEs." *Advances in Neural
Information Processing Systems* 32.

Finlay C, Jacobsen J-H, Nurbekyan L, Oberman AM (2020). "How to train your
neural ODE: the world of Jacobian and kinetic regularization." *Proceedings of
the 37th International Conference on Machine Learning*, 3154-3164.

Massaroli S, Poli M, Park J, Yamashita A, Asama H (2020). "Dissecting neural
ODEs." *Advances in Neural Information Processing Systems* 33, 3952-3963.

Norcliffe A, Bodnar C, Day B, Simidjievski N, Liò P (2020). "On second order
behaviour in augmented neural ODEs." *Advances in Neural Information Processing
Systems* 33, 5911-5921.

Worsham JM, Kalita JK (2025). "A guide to neural ordinary differential
equations: machine learning for data-driven digital engineering." *Digital
Engineering*, 100060. Reference implementation:
<https://github.com/joeworsh/Neural-ODE-Guide>
